

# INSTRUCTIONS

-   This is a group assignment.
-   Submit your answer digitally as two files through Moodle:
    -   An R markdown file (extension **Rmd**). Use the template provided
        to you and provide your answers (both code and text) below each
        question.
    -   An **HTML** file "knitted" by RStudio including all the results and
        plots. More details on how to create these files will be provided
        in class on week 3.
-   Follow the Style Guide (available on Moodle). You can be
    penalized on up to 20% in each question for which you do not follow
    the Style Guide.
-   Questions regarding the assignment should be posted <span class="underline">exclusively</span> on
    the respective discussion forum on Moodle.
-   **Deadline:** 25 February 2024 23:59    


<span class="underline">**Warning:**</span> The detection of <span class="underline">any form of plagiarism</span> in your work
means the assignment will be graded with <span class="underline">ZERO points</span>.

\newpage


# Movie Networks

We are interested in assessing what are the most important movies in
the decade 2010-2019. We will use different strategies to do so.
First, we will load and prepare the data.


## Load and prepare the data

The first step is to load and prepare the movie data. The following
instructions perform some routine data preparation operations. Each
set of instructions is preceded by a comment explaining the procedure.
Run the code below and try to understand each line of code as you
might need to perform some changes.

```{r   }
library(data.table)     # Run once per session
library(ggplot2)        # Run once per session
library(igraph)         # Run once per session

# Load data from file 20200120-imdb_movie_actor.csv (do not forget to
# change your working directory to the folder containing the
# 20200120-imdb_movie_actor.csv file)
dt.movie.actor <- fread("20200120-imdb_movie_actor.csv") 

# Count in how many movies each actor has participated and how many 
# principal actor each movie has
dt.movie.actor[, n_movies := .N, by=actor]
dt.movie.actor[, n_actors := .N, by=list(movie, year)]

# Remove entries in which actors have no name 
dt.movie.actor <- dt.movie.actor[!(actor == "")]

# Save dt.movie.actor. Next time you can simply call the load function (below)
save(dt.movie.actor, file="imdb_movie_actor.RData") 
```

Load the data that you prepared using the instructions below. As mentioned in the
comments, you can start from this line if you have previously saved these data.

```{r   }
# Load previously saved dt.movie.actor. You can
# start in this line if you have previously saved these data.
load("imdb_movie_actor.RData")
```


## Questions (`data.table`) `[7 points]`

This set of questions require that you know how to manipulate a
`data.table`. Answer each of the following questions below. Include
all the code you created/used in your answer.

1.  What is the total amount of movies in the `dt.movie.actor` dataset?
    `[1 point]`

```{r}
# According to the code in the "Load and prepare the data" section, an unique
# movie is defined by the unique combination of the values in columns movie and 
# year, so the same rationale will be applied.
dt.movies.unique <- unique(dt.movie.actor, by = c("movie", "year"))
print("Total amount of movies:")
print(dt.movies.unique[, .N])
# A: Total amount of movies is 368522
```


2.  List the actors from the movie `"Fight Club (1999)"`. List
    the actors from the movie `"Se7en (1995)"`. `[1 point]`

```{r}
# Filter the dataset for the movie we are interested and get the list of actors
l.actors.fight.club <- dt.movie.actor[movie == "Fight Club (1999)", actor]
print("Fight Club Actors:")
print(l.actors.fight.club)
# A: The Fight Club Actors are Brad Pitt (b.1963), Meat Loaf (b.1947),
#    Edward Norton (b.1969) and Zach Grenier (b.1954)

l.actors.se7en <- dt.movie.actor[movie == "Se7en (1995)", actor]
print("Se7en Actors:")
print(l.actors.se7en)
# A: The Se7ven Actors are Brad Pitt (b.1963), Morgan Freeman (b.1937),
#    Kevin Spacey (b.1959) and Andrew Kevin Walker (b.1964)
```

3.  Which actors participated on both movies? Hint: The function
    `intersect` calculates the intersection of two sets. `[1 point]`

```{r}
# Use of intersect to get common actor
actor.intersection <- intersect(l.actors.fight.club, l.actors.se7en)
print("Actors that participated in both 'Fight Club' and 'Se7en")
print(actor.intersection)
# A: The only actor that participated in both movies is Brad Pitt (b.1963)
```

4.  In which movies did Brad Pitt (b.1963) and George Clooney (b.1961) star together?
    `[1 point]`

```{r}
# Find the list of movies for each actor
l.movies.brad.pitt <- dt.movie.actor[actor == "Brad Pitt (b.1963)", movie]
l.movies.george.clooney <- dt.movie.actor[actor == "George Clooney (b.1961)", movie]

# Find the intersection of movies
l.common.movies <- intersect(l.movies.brad.pitt, l.movies.george.clooney)
print("Movies Brad Pitt and George Clooney star together:")
print(l.common.movies)
# A: The movies that both Brad Pitt and George Clooney started together were
#    Ocean's Eleven (2001), Ocean's Twelve (2004), Ocean's Thirteen (2007) and
#    Burn After Reading (2008)
```

5.  Create a table that shows the number of movies released per year.
    This table should include three columns: `year`, `n_movies`, and
    `csum_n_movies`. The first column should contain the year, the
    second the number of movies in that year, and the third, the number
    of movies released since the first year in the data and up to the
    year in that line. Tip: Use the function `cumsum` and check if the
    amount in the last year is the same as the total number of movies
    in question 1. `[1 point]`

```{r}
# Use the data table with unique movies created in question 1
# We order the data table by year and create column n_movies to count the movies
# that released each year
dt.movies.year <- dt.movies.unique[order(year), .(n_movies = .N), by = year]

# Calculate cumulative number of movies using cumsum()
dt.movies.year[, csum_n_movies := cumsum(n_movies)]
print("Data Table:")
print(dt.movies.year)

# Get the value of csum_n_movies in the last year to check with question 1's output.
print("Total number of movies using function cumsum():")
print(dt.movies.year[.N, csum_n_movies])

print("Total number of movies from question 1:")
print(dt.movies.unique[, .N])

# A: Data table with movies per year and cumulative sum stored in dt.movies.year
#    The final value of the cumulative sum and the value obtained in question 1
#    are equal (368,522)
```
We obtain the same number as the total number of movies in question 1: 368522

6.  Which actor/actress has starred in the most movies across all data?
    After (and including) 2000, which year has the most movie
    participations by a single actor/actress? Who is that
    actor/actress? What do these two actors/actresses have in common?
    `[1 point]`

```{r}
# Create a data table with actors and the number of movies for each actor
dt.movies.by.actor <- dt.movie.actor[, .(n_movies = .N), by = actor]

# Order in descending order and get the actor in the first row
actor.most.movies <- dt.movies.by.actor[order(-n_movies)][1, actor]
print("In the whole dataset, the year that has the most movie participations by 
      a single actor/actress is:")
print(actor.most.movies)
# A: Brahmanandam (b.1956)

# Filter movies since 2000
dt.movies.2000 <- dt.movie.actor[year >= 2000]

# Create data table with year, actor and number of movies made
dt.year.actor.2000 <- dt.movies.2000[, .(n_movies = .N), 
                                 by = list(year, actor)]

# Get the year, the actor and the number of movies he/she made
dt.year.most.actor.2000 <- dt.year.actor.2000[n_movies == max(n_movies), ]

print("After (and including) 2000, the year that has the most movie 
      participations by a single actor/actress is:")
print(dt.year.most.actor.2000[, year])
# A: 2001

print("The actor/actress is:")
print(dt.year.most.actor.2000[, actor])
# A: Anil Nagrath

# Trying to find common movies and years between the two actors 
dt.movies.actor.all <- dt.movie.actor[actor == actor.most.movies, ]
dt.movies.actor.2000 <- dt.movie.actor[actor == dt.year.most.actor.2000[, actor], ]

intersection.movies <- intersect(dt.movies.actor.all[, movie], 
                          dt.movies.actor.2000[, movie])
print("Movies Brahmanandam and Anil Nagrath have in common:")
print(intersection.movies)

intersection.years <- intersect(dt.movies.actor.all[, year], 
                          dt.movies.actor.2000[, year])
print("Years when Brahmanandam and Anil Nagrath made movies:")
print(sort(intersection.years))
```
A: We found that the two actors: Brahmanandam and Anil Nagrath did not have any
movies in which they starred together, which was a surprise for us, since both
actors worked in the Indian film industry around the same time, on a great number
of films. One justification for this could be that Brahmanandam primarily worked
in films in Telagu, while Anil Nagrath worked in Hindi films, which would explain
the lack of an intersection between the actors.

7.  Consider only the 10% most popular movies (by votes) in the decade 2010-2019.
    List the top 10 actors that starred in the most movies in the decade.
    Which year(s) has/have the most movie participations by a single actor?
    Hint: you can use the function `quantile` to find how many votes
    does the movie in percentile 90 have. `[1 point]`


```{r}
# Subset to get movies from the decade 2010-2019. For the quantile
# calculation, each movie should counted once
dt.movies.2010.2019 <- unique(dt.movie.actor[year>=2010 & year<=2019], by="movie")

# Get top 10% most popular movies by votes
votes.quantile.90 <- quantile(dt.movies.2010.2019[, votes], 0.9, na.rm = TRUE)
dt.top.movies <- dt.movies.2010.2019[votes >= votes.quantile.90, ]

# Compute the number of participation in top movies and get a list of top 10 actors
dt.actors.top.movies <- dt.movie.actor[movie %in% dt.top.movies[, movie]]
dt.actors.top.movies[, n_movies_decade := .N, by = actor]

dt.unique.actors <- unique(dt.actors.top.movies, by = "actor")
l.top.10.actors <- dt.unique.actors[order(-n_movies_decade)][1:10, actor]
# The 8th, 9th and 10th place may have different actors based on ordering, as 3
# other actors have n_movies_decade = 22
print("Actors that had the most participation in popular movies in 2010-2019:")
print(l.top.10.actors)
# A: Top 10 Actors: Akshay Kumar (b.1967), Nicolas Cage (b.1964), James Franco (b.1978)
#    Robert De Niro (b.1943), Bruce Willis (b.1955), Samuel L. Jackson (b.1948), 
#    Anna Kendrick (b.1985), John Cusack (b.1966), Liam Neeson (b.1952) and
#    Dwayne Johnson (b.1972)

# Get for each actor the number of movies done in each year from top movies
dt.participation <- dt.top.movies[, .(n_movies = .N), by = .(actor, year)]
# Get for each year the most amount of movies done by a single actor
dt.most.participation <- dt.participation[, .(n_movies = max(n_movies)),
                                             by = year]
# Order by descending order and get the year with most participation 
# by a single actor
year.most.participation <- dt.participation[order(-n_movies)][1, year]
print("The year(s) that had the most most movie participations by a single 
      actor/actress:")
print(year.most.participation)
# A: The year with most participation is 2014
```

## Questions (`ggplot2`) `[3 points]`

1.  Plot a histogram with the number of movies per year. Which patterns
    do you observe? Is there anything strange? `[1 point]`

```{r}
# Use dt.movies.year defined in question 5 of group 1
ggplot(dt.movies.year, aes(x = n_movies)) +
  geom_histogram(fill = "grey", color = "black") +
  labs(title = "Distribution of Movies per Year",
       x = "Number of Movies per Year",
       y = "Count of Years") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(breaks = seq(0,13000, by = 1000))

ggplot(dt.movies.year, aes(x = year, y = n_movies)) +
  geom_bar(stat = "identity", fill = "grey", color = "black") +
  labs(title = "Number of Movies per Year",
       x = "Year",
       y = "Number of Movies") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(breaks = seq(1894, 2019, by = 10))
```

In addition to the histogram with the number of movies per year, we also plotted
the evolution of the movies per year in a bar plot.
The most frequent values for the yearly movies number is in the 2000 to 4000 
movies range, with the most frequent value being around 2000 movies. So, the 
pattern is that each year, the amount of movies released are around the range 
mentioned. 
A few years escape this pattern. On hand hand, some years have values way above 
the average, with up to 12000 movie releases. This is likely because overtime, 
the movie industry grew and the number of releases also grew consequently. In 
the bar plot, we can observe the exponential growth in movie production after 
the beginning of the XXI century.
On the other hand, some years have an yearly movie count way below the pattern. 
This is likely because in the oldest years of the data set, the movie industry 
was a lot smaller and the release of movies wasn't nearly as frequent as today.
Naturally, there are many years where the quantity of movies released are in the
single digits. In the bar plot, we can verify this fact, as yearly movie 
production only grew to a considerable size in the 1920s.

2.  Plot a histogram that represents the distribution of number of IMDb
    votes per movie. The x-axis should represent the number of votes
    and the y-axis should represent how many movies have x number of
    votes. Which patterns do you observe? `[1 point]`

```{r}
# Remove movies with NA in the votes column
dt.movies.votes <- na.omit(dt.movies.unique, cols="votes")
# Create Plot
ggplot(dt.movies.votes, aes(x = votes)) +
  geom_histogram(binwidth = 50, fill = "grey", color = "black") +
  labs(title = "Distribution of number of IMDB Votes per Movie",
       x = "Number of Votes",
       y = "Number of Movies") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(limits = c(0, 5000)) + coord_cartesian(ylim = c(0, 50000)) 
```

After removing missing values, the lowest number of votes is 5 votes and the 
maximum number 2,182,052, which justifies why the histogram plotted over a big 
range.When plotting, a limit for the x-axis was set to 5000 for readability 
purposes, so the actual spread of the distribution is more pronounced than 
portrayed.
Regarding the movies with missing data, we adopted the approach of removing them
instead of replacing them by 0, because it would misrepresent the skewness to 
the right of the distribution of the votes. It is worth noting that most movies 
with missing data are older movies, although some recent movies also have NAs.
We can observe that the distribution is positively skewed, with a bigger tail in
the right side of the plot. As we increase the number of votes on the x-axis, 
the number of movies in the y-axis decrease at an exponential rate.
This type of distribution could be explained by network effects in the audience,
since people that enjoy a movie will recommend it to more people, increasing its
popularity exponentially. Since popularity is likely correlated with votes, this
effect would explain the values observed.


3.  Plot a histogram that represents the distribution of the number of
    actors per movie. The x-axis should represent the number of actors
    and the y-axis should represent how many movies have x number of actors.
    Describe your findings. `[1 point]`

```{r}
ggplot(dt.movies.unique, aes(x = n_actors)) +
  geom_histogram(binwidth = 1, fill = "grey", color = "black") +
  labs(title = "Distribution of Number of Actors per Movie",
       x = "Number of Actors",
       y = "Number of Movies") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(limits = c(0, 25))
```

When plotting, bin width was set to 1, since the range of number of actors was 
not very large, and with binwidth = 1 we were able to get a more precise 
representation of the distribution.
The lowest number of actors in a movie is 1 and the highest number is 24. We 
observe that by far the most common number of actors in a movie is 4, with more 
than 200,000 movies, compared to other bins having less than 25,000 movies.
It is important to note that the number of actors in the data set is not
necessarily the total number of actors in the movies. It would actually be odd
for so many movies to have less than 7 10 actors in total. The number of actors
only refers to the number of times there is an entry in the data set for an 
actor starring in a movie, so it is expected to have more complete records for
well-known actors and movies from relatively recent years.

## Questions (`igraph`) `[10 points]`

1.  From this question onwards, and until the end of the assignment,
    focus only on <span class="underline">the actors that participated on the top 50 most
    popular movies from the 2010-2019 decade (by number of votes).</span>
    Load the `igraph` package and create a bipartite graph in which the
    edges correspond to actors&rsquo; participation in movies. How many movie
    participations exist? `[1 point]`
    
```{r}
# Dataframe with movies between 2010 and 2019
dt.movies.2010.2019 <- dt.movie.actor[year >= 2010 & year <= 2019]

# Selecting the top 50 movies from the decade
top.movies <- dt.movies.2010.2019[, .N ,by=list(movie, 
                                                votes)][order(-votes)][1:50, movie]

# Selecting the actors that participate in top 50 movies
actors.in <- dt.movies.2010.2019[movie %in% top.movies]

# Create a bipartite graph with movies and actors 
all.actors <- actors.in[, list(name = unique(actor), type = TRUE)]
all.movies <- actors.in[, list(name = unique(movie), type = FALSE)]
all.vertices <- rbind(all.actors, all.movies)
g <- graph_from_data_frame(actors.in[, list(movie, actor)], 
                           directed = FALSE, vertices = all.vertices)

# Plotting the graph 
plot(g, layout = layout_as_bipartite, vertex.size = 5, vertex.label = NA)

# Number of movie participations
num.edges <- ecount(g)
print("Number of movie participations:")
print(num.edges)
# A: There are 200 participations of actors in the top 50 movies in 
#    the 2010 to 2019 decade
```
2.  Create a graph in which two movies are connected to each other if
    they have <span class="underline">at least one actor in common</span>. Calculate the <span class="underline">degree
    centrality</span> for each of the movies, and remove movies with no
    connections to other movies. <span class="underline">Hint:</span> the function
    `induced.subgraph` allows the creation of graphs with only a subset
    of the vertices. Calculate the following additional centrality
    measures for each of these movies: `[2 points]`
    -   Closeness centrality
    -   Betweenness centrality
    -   Eigenvector centrality

```{r}
# Projection of the bipartite graph with actors as the edges
# & movies as vertex
g.movies.projection <- bipartite_projection(g)$proj1

# Calculating the degree centrality
degree.centrality <- degree(g.movies.projection)

# Checking movies with null degree centrality and excluding them
non.isolates <- which(degree.centrality > 0)
g.proj.filtered <- induced_subgraph(g.movies.projection, non.isolates)

#Plotting the graph
plot(g.proj.filtered, vertex.size=8, vertex.label=NA)

# Creating a datatable, with centrality measures for each film
dt.centrality <- data.table(
  Node = V(g.proj.filtered)$name,
  Degree = degree(g.proj.filtered),
  Closeness = round(closeness(g.proj.filtered), 2),
  Betweenness = betweenness(g.proj.filtered),
  Eigenvector = round(eigen_centrality(g.proj.filtered)$vector, 2)
)

# Printing the datatable with different centralities
print(dt.centrality)
```

3.  For each centrality measure, list the top 20 movies with highest
    centrality. How do you interpret the outcomes? `[2 points]`

```{r}
# For each centrality measure, creating dataframes with the top 20 results
dt.closeness <- dt.centrality[order(-Closeness), .(Node, Closeness)][1:20]
dt.degree <- dt.centrality[order(-Degree), .(Node, Degree)][1:20]
dt.betweenness <- dt.centrality[order(-Betweenness), .(Node, Betweenness)][1:20]
dt.eigenvector <- dt.centrality[order(-Eigenvector), .(Node, Eigenvector)][1:20]

print("Closeness Top 20:")
print(dt.closeness)
print("Degree Top 20:")
print(dt.degree)
print("Betweenness Top 20:")
print(dt.betweenness)
print("Eigenvector Top 20:")
print(dt.eigenvector)
```
The top 20 movies per centrality measure offer insights into their influence 
within the network. Degree centrality reflects direct connections, measuring the
number of common actors with other top 50 movies, and eigenvector centrality 
assesses a movie's influence based on the level of influence of its connected 
movies. This may reflect a own network popular actors, so highly popular movies 
have a high correlation with also highly popular and influential actors. 
Additionally, betweenness centrality, highlighting a node's role in connecting 
others, shows a loose connection with degree centrality, indicating movies with 
higher connectivity also tend to bridge between different parts of the network, 
perhaps by a central netwrok of more popular actors whse influence and 
popularity decreases as one steps away from the center. However, closeness 
centrality, which reflects interaction speed between vertices, shows minimal 
association with the remaining centrality measures.

4.  Calculate the average clustering coefficient for the movies
    network. `[1 point]`

```{r}
# Calculating both local and average clusters
local.cluster <- transitivity(g.proj.filtered, type="local")
average.cluster <- transitivity(g.proj.filtered, type="average")

print("Local Cluster:")
print(local.cluster)
print("Average Cluster:")
print(average.cluster)

# A: The average clustering coefficient is 0,6551873 
```
The clustering coefficient of a node corresponds to the degree to which
its neighbors are linked among themselves as a ratio between the links
their neighbors have and the total amount of links they could have.

A high clustering coefficient suggests that nodes in the network have many
connections to their neighbors - this forms dense clusters. This network 
reflects a real world phenomenon, actors getting hired in high caliber
movies which likely reflects their own caliber as some of the top actors, thus 
this most likely is a higher clustering coefficient than what would be expected 
of an randomized process for links.

5.  Choose one movie you like and plot the movie, their direct
    neighbors and the links among them. What is the clustering
    coefficient of this movie? Which is the actor with most
    participations among these (neighbor) movies, but not having
    participated in the movie itself? `[2 points]`

```{r}
# Selected movie "Avengers: Endgame (2019)
# Getting it's neighbors and plotting a subgraph with only the vertices associated
movie.target <- "Avengers: Endgame (2019)"
neighbors <- neighbors(g.proj.filtered, movie.target)
vertex.indices <- c(which(V(g.proj.filtered)$name == movie.target), 
                    neighbors)
subgraph <- induced_subgraph(g.proj.filtered, vertex.indices)
plot(subgraph, vertex.size = 8, vertex.label = NA)

# Getting the average clustering coefficient
clustering.coefficient <- transitivity(subgraph, type = "average")
print(clustering.coefficient)
# A: The average clustering coefficient is 0.8670996

# Getting the actor/actress with participations in the subset,
# excluding the movie selected
l.subset.movies <- names(neighbors)
dt.a <- actors.in[movie %in% l.subset.movies, ]
dt.actors.target <- actors.in[movie %in% "Avengers: Endgame (2019)", ]
dt.unique.movies <- dt.a[!(dt.a$actor %in% dt.actors.target$actor)]

# Getting the actor/actress with the most partipations
dt.top.actor <- dt.unique.movies[, n := .N, by = actor][order(-n)][1]
dt.top.actor[, list(actor,n)]
#A: Scarlett Johansson with 3 movies
```
By selecting a movie and only it's neighbors, that is the movies that
directly share one actor between each other, we get a select cluster of
movies. In this case, the average clustering coefficient for this network
is superior to the one seen before: by selecting only a movie and its
neighbors there is a higher density of ties, and, as characteristic in
real-world networks, it scores a high coefficient - it is not random that there is a pattern for the same actors in the same circle of movies.

6.Plot the degree distribution of the movies. How do you compare them with
the degree distribution of a random graph? What can be plausible
explanations for the observed differences? `[2 points]`


```{r}
# Histogram with degree distribution of the top 50 movies
ggplot() + geom_histogram(aes(x = degree(g.proj.filtered)), binwidth = 1, 
                          fill = "grey", color = "black") +  
  labs(title = "Distribution of Number of Actors per Movie",
       x = "Degree",
       y = "Count") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Degree Distribution of Movies")
```

The observed distribution demonstrates a pronounced right skew, indicating
that movies generally exhibit lower levels of degree centrality. In
contrast to the anticipated normal distribution of a random graph,
real-world social networks often display fat tails, characterized by an
abundance of nodes with minimal links as well as a surplus of highly
connected nodes. This distribution aligns closely with a scale-free or
power-law distribution, diverging from the typical random graph model.
