

# INSTRUCTIONS

-   This is a group assignment.
-   Submit your answer digitally as two files through Moodle:
    -   An R markdown file (extension **Rmd**). Use the template provided
        to you and provide your answers (both code and text) below each
        question.
    -   An **HTML** file "knitted" by RStudio including all the results and
        plots. More details on how to create these files will be provided
        in class on week 3.
-   Follow the Style Guide (available on Moodle). You can be
    penalized on up to 20% in each question for which you do not follow
    the Style Guide.
-   Questions regarding the assignment should be posted <span class="underline">exclusively</span> on
    the respective discussion forum on Moodle.
-   **Deadline:** 25 February 2024 23:59    


<span class="underline">**Warning:**</span> The detection of <span class="underline">any form of plagiarism</span> in your work
means the assignment will be graded with <span class="underline">ZERO points</span>.

\newpage


# Movie Networks

We are interested in assessing what are the most important movies in
the decade 2010-2019. We will use different strategies to do so.
First, we will load and prepare the data.


## Load and prepare the data

The first step is to load and prepare the movie data. The following
instructions perform some routine data preparation operations. Each
set of instructions is preceded by a comment explaining the procedure.
Run the code below and try to understand each line of code as you
might need to perform some changes.

```{r   }
library(data.table)     # Run once per session
library(ggplot2)        # Run once per session

# Load data from file 20200120-imdb_movie_actor.csv (do not forget to
# change your working directory to the folder containing the
# 20200120-imdb_movie_actor.csv file)
dt.movie.actor <- fread("20200120-imdb_movie_actor.csv")

# Count in how many movies each actor has participated and how many 
# principal actor each movie has
dt.movie.actor[, n_movies := .N, by=actor]
dt.movie.actor[, n_actors := .N, by=list(movie, year)]

# Remove entries in which actors have no name 
dt.movie.actor <- dt.movie.actor[!(actor == "")]

# Save dt.movie.actor. Next time you can simply call the load function (below)
save(dt.movie.actor, file="imdb_movie_actor.RData")
```

Load the data that you prepared using the instructions below. As mentioned in the
comments, you can start from this line if you have previously saved these data.

```{r   }
# Load previously saved dt.movie.actor. You can
# start in this line if you have previously saved these data.
load("imdb_movie_actor.RData")
```


## Questions (`data.table`) `[7 points]`

This set of questions require that you know how to manipulate a
`data.table`. Answer each of the following questions below. Include
all the code you created/used in your answer.

1.  What is the total amount of movies in the `dt.movie.actor` dataset?
    `[1 point]`

```{r}
# According to the code in the "Load and prepare the data" section, an unique
# movie is defined by the unique combination of the values in columns movie and 
# year, so the same rationale will be applied.
dt.movie.unique <- unique(dt.movie.actor, by = c("movie", "year"))
print(dt.movie.unique[, .N])
#OR
dt.movie.unique <- dt.movie.actor[, .N, by = list(movie, year)]
print(nrow(dt.movie.unique))
```


2.  List the actors from the movie `"Fight Club (1999)"`. List
    the actors from the movie `"Se7en (1995)"`. `[1 point]`

```{r}
l.actor.fight.club <- dt.movie.actor[movie == "Fight Club (1999)", actor]
print(l.actor.fight.club)
l.actor.se7en <- dt.movie.actor[movie == "Se7en (1995)", actor]
print(l.actor.se7en)
```

3.  Which actors participated on both movies? Hint: The function
    `intersect` calculates the intersection of two sets. `[1 point]`

```{r}
actor.intersection <- intersect(l.actor.fight.club, l.actor.se7en)
print(actor.intersection)
```

4.  In which movies did Brad Pitt (b.1963) and George Clooney (b.1961) star together?
    `[1 point]`

```{r}
l.movie.brad.pitt <- dt.movie.actor[actor == "Brad Pitt (b.1963)", movie]
l.movie.george.clooney <- dt.movie.actor[actor == "George Clooney (b.1961)", movie]
l.common.movies <- intersect(l.movie.brad.pitt, l.movie.george.clooney)
print(l.common.movies)
```

5.  Create a table that shows the number of movies released per year.
    This table should include three columns: `year`, `n_movies`, and
    `csum_n_movies`. The first column should contain the year, the
    second the number of movies in that year, and the third, the number
    of movies released since the first year in the data and up to the
    year in that line. Tip: Use the function `cumsum` and check if the
    amount in the last year is the same as the total number of movies
    in question 1. `[1 point]`

```{r}
dt.movie.year <- dt.movie.unique[order(year), .(n_movies = .N), by = year]
dt.movie.year[, csum_n_movies := cumsum(n_movies)]
print(dt.movie.year)
print(dt.movie.year[.N, csum_n_movies])
```

6.  Which actor/actress has starred in the most movies across all data?
    After (and including) 2000, which year has the most movie
    participations by a single actor/actress? Who is that
    actor/actress? What do these two actors/actresses have in common?
    `[1 point]`

```{r}
actor.most.movies.all <- unique(dt.movie.actor[n_movies == max(n_movies), actor])
print(actor.most.movies.all)


dt.year.actor.2000 <- dt.movie.actor[year >= 2000, .(n_movies = .N), 
                                     by = list(year,actor)]
dt.year.most.actor.2000 <- dt.year.actor.2000[n_movies == max(n_movies), ]
print(dt.year.most.actor.2000[, year])
print(dt.year.most.actor.2000[, actor])

l.movie.actor.all <- dt.movie.actor[actor == actor.most.movies.all, movie]
l.movie.actor.2000 <- dt.movie.actor[actor == dt.year.most.actor.2000[, actor], movie]
print(intersect(l.movie.actor.all, l.movie.actor.2000))
```

7.  Consider only the 10% most popular movies (by votes) in the decade 2010-2019.
    List the top 10 actors that starred in the most movies in the decade.
    Which year(s) has/have the most movie participations by a single actor?
    Hint: you can use the function `quantile` to find how many votes
    does the movie in percentile 90 have. `[1 point]`


```{r}
# Movies with no votes will be ignored in the quantile calculation
dt.movie.2010.2019 <- dt.movie.actor[year>=2010 & year<=2019, .(votes = votes[1]),
                                     by = movie]
votes.quantile.90 <- quantile(dt.movie.2010.2019[, votes], 0.9, na.rm = TRUE)
dt.movie.2010.2019.popular <- dt.movie.2010.2019[votes >= votes.quantile.90, ]

dt.movie.actor.filtered <- dt.movie.actor[movie %in% dt.movie.2010.2019.popular, ]
dt.movie.actor.filtered[, n_movies_decade := .N, by = actor]
l.top.10.actors <- unique(dt.movie.actor.filtered, 
                           by = "actor")[order(-n_movies_decade)][1:10, actor]
print(l.top.10.actors)

dt.movie.actor.filtered[, n_movies_year := .N, by = list(year, actor)]
year.most.actor.2000.2019 <- unique(dt.movie.actor.filtered[
                                    n_movies_year == max(n_movies_year), year])
print(year.most.actor.2000.2019)
```

## Questions (`ggplot2`) `[3 points]`

1.  Plot a histogram with the number of movies per year. Which patterns
    do you observe? Is there anything strange? `[1 point]`

```{r}
ggplot(dt.movie.year, aes(x = n_movies)) + geom_histogram()

# Most years have a value of n_movies (variable defined for the yearly movie 
# count) in the 0 to 4000 movies range, with most frequent n_movies value being 
# around 2000 movies, and with years having a greater amount of n_movies, up to 
# 12000 movies. 
# One thing that is strange is that there are 10 years with a n_movies value near
# 0. This is likely related to older years in the data, where both the number of
# movies released and the information about them would be scarcer than the average
# year. By using the View() function, we can confirm that these low values belong
# to years in the end of the XIX / beginning of the XX centuries.
```

2.  Plot a histogram that represents the distribution of number of IMDb
    votes per movie. The x-axis should represent the number of votes
    and the y-axis should represent how many movies have x number of
    votes. Which patterns do you observe? `[1 point]`

```{r}
ggplot(dt.movie.unique, aes(x = votes)) + geom_histogram()

# ggplot plot the histogram with a range on the x axis from 0 to around 2 million
# votes, where most movies are on the lower end of the histogram, while very few
# ones have extremely high number of votes. This can be a factor of old age and
# lack of historical relevance of the movies, as a lot of old movies that are not
# very famous have votes in the thousands or below. In a lot of cases, there aren't
# even any votes, as observable in the warning message give by ggplot in the output,
# where it is indicated that 153697 rows were removed from the plot because of NAs.
```

3.  Plot a histogram that represents the distribution of the number of
    actors per movie. The x-axis should represent the number of actors
    and the y-axis should represent how many movies have x number of actors.
    Describe your findings. `[1 point]`

```{r}
ggplot(dt.movie.unique, aes(x = n_actors)) + geom_histogram()

# From the histogram, we can observe that the most frequent value is 4, and that
# most values are concentrated between 0 and 10. There is a very high imbalance 
# between the frequency of the value 4 and the others, with 4 having a count higher
# than 200,000, while the second most popular value, 5, reaches just over 25000.
# Also, ggplot plots over a range in the x-axis from 0 to just over 25, meaning
# that there are a few movies that have a number of actors in the data way higher
# than the most frequent and average values. It is important to note, however,
# that the n_actors number is not necessarily the total number of actors in the
# movie, but instead is the number of times an entry for an actor starring in that
# movie is recorded in the data set. So normally, we can expect older and/or less
# relevant movies to have incomplete information about the actors starred in them.
```

## Questions (`igraph`) `[10 points]`

1.  From this question onwards, and until the end of the assignment,
    focus only on <span class="underline">the actors that participated on the top 50 most
    popular movies from the 2010-2019 decade (by number of votes).</span>
    Load the `igraph` package and create a bipartite graph in which the
    edges correspond to actors&rsquo; participation in movies. How many movie
    participations exist? `[1 point]`
    
```{r}

dt.movies.2010.2019 <- dt.movie.actor[year>=2010 & year<=2019]
top.movies <- dt.movies.2010.2019[, .N ,by=list(movie,votes)][order(-votes)][1:50, movie]
actors.in <- dt.movies.2010.2019[movie %in% top.movies]
all.actors <- actors.in[, list(name=unique(actor), type=TRUE)]
all.movies <- actors.in[, list(name=unique(movie), type=FALSE)]
all.vertices <- rbind(all.actors, all.movies)
library(igraph)
g <- graph_from_data_frame(actors.in[, list(movie, actor)], directed=FALSE, vertices=all.vertices)
plot(g, layout = layout_as_bipartite, vertex.size = 5, vertex.label=NA)
num_edges <- ecount(g)
print(num_edges)
```


2.  Create a graph in which two movies are connected to each other if
    they have <span class="underline">at least one actor in common</span>. Calculate the <span class="underline">degree
    centrality</span> for each of the movies, and remove movies with no
    connections to other movies. <span class="underline">Hint:</span> the function
    `induced.subgraph` allows the creation of graphs with only a subset
    of the vertices. Calculate the following additional centrality
    measures for each of these movies: `[2 points]`
    -   Closeness centrality
    -   Betweenness centrality
    -   Eigenvector centrality

```{r}
# <answer here>

g.movies <- bipartite.projection(g)$proj1
degree_centrality <- degree(g.movies)
isolates <- which(degree_centrality == 0)
g.movies <- delete_vertices(g.movies, isolates)
plot(g.movies, vertex.label = NA)

centrality_data <- data.table(
  Node = V(g.movies)$name,
  Degree = degree(g.movies),
  Closeness = round(closeness(g.movies), 2),
  Betweenness = betweenness(g.movies),
  Eigenvector = round(eigen_centrality(g.movies)$vector, 2)
)
print(centrality_data)

```

3.  For each centrality measure, list the top 20 movies with highest
    centrality. How do you interpret the outcomes? `[2 points]`

```{r}
# <answer here>
top_closeness <- centrality_data[order(-Closeness), .(Node, Closeness)][1:20]
top_degree <- centrality_data[order(-Degree), .(Node, Degree)][1:20]
top_betweenness <- centrality_data[order(-Betweenness), .(Node, Betweenness)][1:20]
top_eigenvector <- centrality_data[order(-Eigenvector), .(Node, Eigenvector)][1:20]

print(top_closeness)
print(top_degree)
print(top_betweenness)
print(top_eigenvector)
```

4.  Calculate the average clustering coefficient for the movies
    network. `[1 point]`

```{r}
# <answer here>
local.cluster <- transitivity(g.movies, type="local")
average.cluster <- transitivity(g.movies, type="average")

print(local.cluster)
print(average.cluster)
```

5.  Choose one movie you like and plot the movie, their direct
    neighbors and the links among them. What is the clustering
    coefficient of this movie? Which is the actor with most
    participations among these (neighbor) movies, but not having
    participated in the movie itself? `[2 points]`

```{r}
# <answer here>
movie_target <- "Avengers: Endgame (2019)"
neighbors <- neighbors(g.movies, movie_target)
vertex_indices <- c(which(V(g.movies)$name == movie_target), neighbors)
subgraph <- induced_subgraph(g.movies, vertex_indices)
plot(subgraph)


clusteringCoefficient <- transitivity(subgraph, type = "local")
print(clusteringCoefficient)


subset_movies <- names(neighbors)
a <- actors.in[movie %in% subset_movies, ]
actors.target <- actors.in[movie %in% "Avengers: Endgame (2019)", ]
unique_movies <- a[!(a$actor %in% actors.target$actor)]
top.actor <- unique_movies[,n := .N,by = actor][order(-n)][1]
top.actor[,list(actor,n)]

```

6.  How do you compare them
    with the degree distribution of a random graph? What can be
    plausible explanations for the observed differences? `[2 points]`


```{r}
# <answer here>
ggplot() + geom_histogram(aes(x = degree(g.movies)), binwidth = 1)
```
